---
title: "Video Game Sales Data Analysis"
author: "Rohan Gupta, Kayleigh Derwin, Will Bolger"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

# Introduction

This document contains analyses and visualizations for video games sales accross different regions of the world.

# Setup

```{r}
# Check and install necessary libraries
packages <- c("ggplot2", "dplyr", "knitr", "kableExtra", "readr", "tidyr")

for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

# Uncomment the line bellow to change the working directory
# setwd("< Path to your working directory >")
```


## This report aims to answer 3 questions about the video games sales using the data

Q1. How are the various publishers performing in different regions of the world when it comes to sale of their games?

Q2. Does the genre of the game affect its sales? What genre of game is most popular among each region of the world?

Q3. Are there any stand out games per year that made a publisher really successfull or unsuccessful?


## Understanding the Database

For this project we are using a CSV file as our source of data. This file named "PS4_GamesSales.csv" has been sources from Kaggle.

The data has 1,301 entries that specify the name, year of release, genre, publisher, and sales accross different continents of the world for video games until the end of the year 2018.

Loading the data
```{r}
data <- read.csv("PS4_GamesSales.csv")
```

Displaying the top ten rows of the data to provide a glimps of the data

```{r}
# Extract the top five entries from the dataset
top_five_entries <- head(data, 5)

# Create a nicely formatted table using kable and kableExtra
kable_styled <- kable(top_five_entries, format = "html", caption = "Top Five PS4 Game Sales", align = 'l') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  add_header_above(c(" " = 1, "Game Data" = 3, "Regional Sales in Millions" = 5)) %>%  
  column_spec(1, bold = TRUE, border_right = TRUE, extra_css = "border-bottom: 1px solid #dddddd;") %>%
  column_spec(2:9, extra_css = "border-right: 1px solid #dddddd; border-bottom: 1px solid #dddddd;") %>%
  scroll_box(width = "100%", height = "300px")

# Print the styled table
kable_styled
```


## Question 1


#### How are the various publishers performing in different regions of the world when it comes to sale of their games?

The following chart displays the total sales of the given publisher according to each region and annually
```{r}
# Relace the value for variable provided with the name of the desired publisher
publisher <- "EA Sports"

# Filter data for the specified publisher "Rockstar Games"
publisher_data <- data %>%
  filter(Publisher == publisher)

# Summarize sales by year and region
annual_sales_by_region <- publisher_data %>%
  group_by(Year) %>%
  summarise(
    Sales_North_America = sum(North_America, na.rm = TRUE),
    Sales_Europe = sum(Europe, na.rm = TRUE),
    Sales_Japan = sum(Japan, na.rm = TRUE),
    Sales_Rest_of_World = sum(Rest_of_World, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = starts_with("Sales"),
    names_to = "Region",
    values_to = "Sales"
  )

# Create the stacked bar chart
ggplot(annual_sales_by_region, aes(x = as.factor(Year), y = Sales, fill = Region)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(title = paste("Annual Sales Distribution by Region for", publisher),
       x = "Year",
       y = "Sales (Millions)",
       fill = "Region") +
  scale_y_continuous(labels = scales::comma)

```



## Question 2

#### Does the genre of the game affect its sales? What genre of game is most popular among each region of the world?


First finding the total sales of games by genre overall the world

```{r}
# Aggregate global sales by genre
genre_sales_Global <- data %>%
  group_by(Genre) %>%
  summarize(Total_Global_Sales = sum(Global, na.rm = TRUE)) %>%
  arrange(desc(Total_Global_Sales))

# Create a bar chart comparing genres based on global sales
ggplot(genre_sales_Global, aes(x = Genre, y = Total_Global_Sales, fill = Genre)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Global Sales by Genre", x = "Genre", y = "Total Global Sales (Millions)")
```

This graphs showed that the top three genres of games with most sales are:

###### 1. Shooter
###### 2. Action
###### 3. Sports


#### Sales in Different Regions of the world

The following set of bar graphs display the total sales of games by genre in each region of the world
```{r plot, fig.width=10, fig.height=6, out.width='100%'}
# Aggregate sales by genre for each region
genre_sales <- data %>%
  group_by(Genre) %>%
  summarize(
    Sales_North_America = sum(North_America, na.rm = TRUE),
    Sales_Europe = sum(Europe, na.rm = TRUE),
    Sales_Japan = sum(Japan, na.rm = TRUE),
    Sales_Rest_of_World = sum(Rest_of_World, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = starts_with("Sales"),
    names_to = "Region",
    values_to = "Sales"
  )

# Create a faceted bar chart for all regions with modified x-axis
ggplot(genre_sales, aes(x = Genre, y = Sales, fill = Genre)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Region, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(title = "Sales by Genre across Different Regions", x = "", y = "Sales (Millions)") +
  guides(fill = guide_legend(title = "Genre"))

```


For a better understanding of the top performing genres, the followning table displays the top 3 performing genres in each region of the world
```{r}
# Function to get top 3 genres in a specific region and reshape for table
get_top_genres_reshaped <- function(data, sales_column, region_name) {
  top_genres <- data %>%
    group_by(Genre) %>%
    summarise(Total_Sales = sum(!!sym(sales_column), na.rm = TRUE)) %>%
    arrange(desc(Total_Sales)) %>%
    slice_head(n = 3) %>%
    mutate(Region = region_name,
           Rank = paste0("Rank_", row_number()),  # Create a ranking label for merging
           Genre_Sales = paste(Genre, "(", Total_Sales, "M)")) %>%  # Combine genre and sales
    select(Region, Rank, Genre_Sales) %>%
    pivot_wider(names_from = Rank, values_from = Genre_Sales)  # Reshape data to wide format
  return(top_genres)
}

# Get top 3 genres for each region, reshaped
top_genres_North_America <- get_top_genres_reshaped(data, "North_America", "North America")
top_genres_Europe <- get_top_genres_reshaped(data, "Europe", "Europe")
top_genres_Japan <- get_top_genres_reshaped(data, "Japan", "Japan")
top_genres_Rest_of_World <- get_top_genres_reshaped(data, "Rest_of_World", "Rest of the World")
top_genres_Global <- get_top_genres_reshaped(data, "Global", "Global")

# Combine all regions into one dataframe
top_genres_all_regions <- bind_rows(
  top_genres_North_America,
  top_genres_Europe,
  top_genres_Japan,
  top_genres_Rest_of_World,
  top_genres_Global
)

# Display the table using kable and kableExtra for enhanced formatting
table_output <- kable(top_genres_all_regions, format = "html", caption = "Top 3 Genres by Sales in Each Region") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  row_spec(0, bold = TRUE, background = "#4D5B69", color = "white") %>%  # Style header row
  column_spec(1, width = "20%", border_right = TRUE) %>%  # Style first column with a border
  column_spec(2, width = "25%", border_right = TRUE) %>%  # Style second column with a border
  column_spec(3, width = "25%", border_right = TRUE) %>%  # Style third column with a border
  column_spec(4, width = "25%", border_right = TRUE) %>%  # Style fourth column with a border
  scroll_box(width = "100%", height = "300px")  # Adds a scroll box if the table is too long

# Print the styled table
table_output
```

#### Fun fact observed in the data

An intriguing observation from our analysis of video game genre sales across different global regions reveals a distinct pattern. While the preferences for game genres in North America, Europe, and the Rest of the World—including the global market—tend to align closely across various rankings, Japan consistently demonstrates unique preferences that set it apart. This distinct difference in genre popularity not only highlights the top three genres but extends throughout the entire spectrum of game genres. This phenomenon underscores Japan's unique gaming culture and market dynamics, which are markedly different from other regions, reflecting broader cultural and consumer taste variations.


## Question 3

#### Are there any stand out games per year that made a publisher really successfull or unsuccessful?

This question is slightly more complicated as compared to the other 2 questions this report deals with. Thus to answer this question we would follow a step wise program discribed below:

Step 1: Find out the highest grossing video game from all publishers.

Step 2: Find the mean of the global sales for all video games that came before and after the highest grossing video game. Both before and after mean should be stored seperatly.

Step 3: Compare the difference between the before and after mean. Also compare the after mean with the sales of the highest grossing game. If there is a huge positive difference between the before sales mean and after sales mean and the after sales mean and sales of the highest grossing game are close to one and other. This would indicate the highpothesis was correct, and stand out games do help make publishers more successful.

<br>

#### Step 1:
```{r}
# Find the highest grossing game globally from each publisher
highest_grossing_games <- data %>%
  group_by(Publisher) %>%
  filter(Global == max(Global)) %>%
  summarise(Game = first(Game), 
            Year = first(Year), 
            Global_Sales = max(Global)) %>%
  ungroup() %>%
  arrange(desc(Global_Sales))

# Create an enhanced table using kable and kableExtra
table_output <- kable(highest_grossing_games, format = "html", 
                      caption = "Highest Grossing Games by Publisher Globally") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, width = "20%") %>%
  column_spec(3, width = "10%") %>%
  column_spec(4, width = "10%", color = "blue") %>%
  scroll_box(width = "100%", height = "400px")

# Print the styled table
table_output
```

<br>

#### Step 2:
```{r}
# Extend the data frame to include more metrics
extended_data <- data %>%
  group_by(Publisher) %>%
  mutate(Highest_Global_Sales = max(Global)) %>%
  filter(Global == Highest_Global_Sales) %>%
  summarise(
    Game = first(Game),
    Year = first(Year),
    Global_Sales = first(Global),
    Mean_Before = mean(data$Global[data$Publisher == first(Publisher) & data$Year < first(Year)], na.rm = TRUE),
    Mean_After = mean(data$Global[data$Publisher == first(Publisher) & data$Year > first(Year)], na.rm = TRUE)
  ) %>%
  arrange(desc(Global_Sales))

# Create an enhanced table using kable and kableExtra
table_output <- kable(extended_data, format = "html", 
                      caption = "Highest Grossing Games by Publisher Globally with Mean Sales Before and After") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, width = "15%") %>%
  column_spec(3, width = "10%") %>%
  column_spec(4, width = "10%", color = "blue") %>%
  column_spec(5, width = "15%", color = "red") %>%
  column_spec(6, width = "15%", color = "green") %>%
  scroll_box(width = "100%", height = "400px")

# Print the styled table
table_output
```

<br>

#### Step 3:
```{r}
# Calculate mean sales before and after the highest grossing game for each publisher
mean_sales_data <- data %>%
  group_by(Publisher) %>%
  summarize(
    Highest_Global_Sales = max(Global),
    Year_of_Highest_Sales = Year[which.max(Global)],
    Mean_Before = mean(Global[Year < Year_of_Highest_Sales], na.rm = TRUE),
    Mean_After = mean(Global[Year > Year_of_Highest_Sales], na.rm = TRUE)
  ) %>%
  ungroup()

# Calculate the overall mean of these averages
overall_mean_before = mean(mean_sales_data$Mean_Before, na.rm = TRUE)
overall_mean_after = mean(mean_sales_data$Mean_After, na.rm = TRUE)

# Print the results
cat("Overall Mean of Average Sales Before the Best Seller: ", overall_mean_before, "\n")
cat("Overall Mean of Average Sales After the Best Seller: ", overall_mean_after, "\n")
```

<br>

Here it can be observed that the average sales after the release of the best selling video game is lower than the average sales before the best selling video game. This indicatest that the prediction was wrong.


## Conclusion

The performance of each publisher can be noted in the charts associated with questions 1

Overall the top 3 genres for video game sales are are shooter, action and sports. However, this trend changes for Japan.

The best selling games for any publisher generally lowers their sales by approximately 500,000 on average.






